\documentclass[11pt]{article}

\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage[table,dvipsnames]{xcolor}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,shapes.geometric,fit,backgrounds,calc}
\usepackage[most]{tcolorbox}
\usepackage{float}
\usepackage[hidelinks]{hyperref}

\emergencystretch=3em  % absorb the occasional unbreakable prose line

% Plain-language callout: a non-technical summary readers can rely on without
% reading the math around it.
\newtcolorbox{plain}{
  enhanced, breakable,
  colback=Sepia!7, colframe=Sepia!55, boxrule=0.4pt, arc=2pt,
  left=7pt, right=7pt, top=4pt, bottom=4pt,
  fontupper=\small,
  coltitle=Sepia!72!black, fonttitle=\bfseries\sffamily\footnotesize,
  title={IN PLAIN TERMS}
}

% Notation callout: same framing, different color/heading from the plain box.
\newtcolorbox{notation}{
  enhanced, breakable,
  colback=infoblue!6, colframe=infoblue!55, boxrule=0.4pt, arc=2pt,
  left=7pt, right=7pt, top=4pt, bottom=4pt,
  fontupper=\small,
  coltitle=infoblue!80!black, fonttitle=\bfseries\sffamily\footnotesize,
  title={WHAT THE SYMBOLS MEAN}
}

% Definition callout: for normative-ish local definitions used by the method.
\newtcolorbox{definitionbox}{
  enhanced, breakable,
  colback=infoblue!4, colframe=infoblue!45, boxrule=0.4pt, arc=2pt,
  left=7pt, right=7pt, top=4pt, bottom=4pt,
  fontupper=\small,
  coltitle=infoblue!80!black, fonttitle=\bfseries\sffamily\footnotesize,
  title={DEFINITION}
}

% --- palette ---
\definecolor{crit}{HTML}{C0392B}
\definecolor{high}{HTML}{E67E22}
\definecolor{med}{HTML}{F1C40F}
\definecolor{low}{HTML}{27AE60}
\definecolor{infoblue}{HTML}{2C3E50}
\definecolor{tagbg}{HTML}{ECF0F1}
\definecolor{rule}{HTML}{BDC3C7}

\newcommand{\tagn}[1]{{\footnotesize\ttfamily #1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\rfckw}[1]{\textsc{#1}}
\newcommand{\vdrcite}{\href{https://github.com/FedRAMP/2026-markdown/blob/2a75592f0e1f86a48b2888d29af6c570c419a7a5/providers/rev5/rules/vulnerability-detection-and-response.md}{FedRAMP/2026-markdown@2a75592}}
\newcommand{\vercite}{\href{https://github.com/FedRAMP/2026-markdown/blob/2a75592f0e1f86a48b2888d29af6c570c419a7a5/providers/rev5/rules/vulnerability-evaluation-and-reporting.md}{FedRAMP/2026-markdown@2a75592}}
\newcommand{\defcite}{\href{https://github.com/FedRAMP/2026-markdown/blob/2a75592f0e1f86a48b2888d29af6c570c419a7a5/definitions.md}{FedRAMP/2026-markdown@2a75592}}

\title{\bfseries A Deterministic, CVSS--Environmental Method for\\
FedRAMP Rev5 VDR/VER Vulnerability Prioritization}
\author{Matthew Venne \\ \small Chief Technology Officer, stackArmor}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
The FedRAMP Vulnerability Detection and Response (VDR) and Vulnerability
Evaluation and Reporting (VER) rules define \emph{Potential Agency Impact} (PAIN,
levels N1--N5) and a remediation--timeframe matrix, together with evaluation rules
for likely exploitability and internet reachability. FedRAMP deliberately does
\emph{not} prescribe how to derive a PAIN level from an individual vulnerability on
a specific asset. This memo argues that the missing classifier already exists, in
standardized form, as the \textbf{Environmental metric group of CVSS}: the
Confidentiality, Integrity and Availability \emph{Requirements} (CR/IR/AR) and the
Modified Impact Sub-Score were designed precisely to re-weight a vulnerability's
impact by what is at stake in a given deployment. We present a closed-form,
reproducible mapping from FedRAMP's requirements onto CVSS Environmental metrics,
derive PAIN and the remediation deadline from it, and give fully worked
calculations. We also present an \emph{example} mechanism --- ``asset archetypes''
--- for assigning CR/IR/AR systematically, while emphasizing that each Cloud Service
Provider (CSP) \rfckw{should} derive its own assignment from its system
categorization. A reference architecture and a sample scan complete the memo.
\end{abstract}

\tableofcontents

\section{Status of This Memo and Terminology}
This document is informational. It does not itself define a FedRAMP requirement; it
specifies a standardized \emph{method} for satisfying the existing VDR/VER
requirements.\footnote{The VDR and VER rules launched on 2026-06-24 as part of the
FedRAMP Consolidated Rules for 2026 and apply to both FedRAMP 20x and Rev5
offerings. This memo retains its original Rev5-oriented title for link stability;
the method applies unchanged under both.}

The key words \rfckw{must}, \rfckw{must not}, \rfckw{should}, \rfckw{should not},
and \rfckw{may} are to be interpreted as described in RFC~2119.

\begin{description}[leftmargin=1.5em,style=nextline]
\item[PAIN] Potential Agency Impact, the FedRAMP N1--N5 rating of a finding.
\item[CR/IR/AR] The CVSS Environmental Confidentiality / Integrity / Availability
  \emph{Requirements} of an asset.
\item[ISC] Impact Sub-Score (here, the CVSS \emph{Modified} Impact Sub-Score).
\item[LEV / NLEV] Likely Exploitable Vulnerability / Not Likely Exploitable
  Vulnerability (a VER axis).
\item[IRV / NIRV] Internet-Reachable Vulnerability / Not Internet-reachable
  Vulnerability (a VER axis).
\item[VEX] Vulnerability Exploitability eXchange, a machine-readable assertion about
  whether a finding affects a specific product or deployment.
\item[Class] The provider Certification Class (A--D) selecting a remediation table.
\end{description}

\section{Motivation: the gap between VDR/VER and the scanner}
A vulnerability scanner emits, per finding, a CVE identifier, a qualitative
severity, and (often) a CVSS Base vector. FedRAMP Rev5 VDR/VER, by contrast, asks
the provider to answer a different question: \emph{what is the potential impact on
the agency if this vulnerability is exploited on this asset, and how quickly must
it therefore be remediated?} The two are not the same. A ``Low''-rated CVE on a
crown-jewel datastore can carry more agency impact than a ``High''-rated CVE on a
throwaway sandbox.

FedRAMP supplies the \emph{output} vocabulary (the PAIN buckets and the remediation
matrix) and the \emph{evaluation axes} (likely-exploitable, internet-reachable) but
leaves the \emph{classifier} --- the function from (CVE, asset) to PAIN --- to the
provider. An ad-hoc or analyst-by-analyst classifier is neither reproducible nor
defensible. This memo supplies a deterministic one.

\section{Key insight: FedRAMP's requirements are CVSS Environmental metrics}
The central claim of this memo is that FedRAMP's notion of ``impact on the agency''
maps, almost one-to-one, onto metrics that already exist in the CVSS v3.1
specification's \emph{Environmental} group. The Environmental group exists for
exactly this purpose: to let the consuming organization re-score a vulnerability
according to the criticality of the affected asset in \emph{its} environment.

\begin{table}[H]
\centering
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.44\textwidth} >{\raggedright\arraybackslash}p{0.44\textwidth}}
\toprule
\textbf{FedRAMP VDR/VER concept} & \textbf{CVSS mechanism that satisfies it} \\
\midrule
Asset data sensitivity / FIPS-199 C-I-A categorization & Security Requirements
\code{CR}, \code{IR}, \code{AR} \\
Magnitude of potential agency impact & Modified Impact Sub-Score (\code{ISC}) \\
Likely exploitable (VER) & \code{LEV}: EPSS threshold / CISA KEV catalog membership \\
Internet reachable (VER) & \code{IRV}: per-finding payload reachability (companion method) \\
Single- vs.\ multi-agency blast radius & scope multiplier on the PAIN tier \\
Provider Certification Class & selector of the remediation deadline table \\
\bottomrule
\end{tabular}
\caption{FedRAMP requirements fall directly onto CVSS Environmental metrics.}
\end{table}

The consequence is that a conforming implementation need not invent a bespoke
risk algebra. It need only (a) assign each asset its CR/IR/AR, and (b) evaluate the
two VER axes. The arithmetic that turns those inputs into a defensible impact
magnitude is the published CVSS Environmental formula.

\section{The model}
We define PAIN as a function of a \emph{severity} term and a \emph{scope} term:
\[
\text{PAIN}(\text{finding}) \;=\; f\big(\,\text{SEVERITY},\ \text{SCOPE}\,\big).
\]
\textbf{SEVERITY} answers ``how badly does this vulnerability's impact land on
\emph{this} asset?'' --- the CVE's impact metrics re-weighted by the asset's
CR/IR/AR. \textbf{SCOPE} answers ``how many agencies' data does the asset hold?''
--- single (1) or multiple ($>1$). SEVERITY is mapped to a FedRAMP customer-effect
word (Minimal, Narrow, Disruptive, Debilitating); the word and scope together
select the N-level.

\section{Two threat inputs: EPSS and the CISA KEV catalog}
\label{sec:enrichment}
\begin{plain}
Beyond the CVE's own CVSS vector, the method pulls in exactly two facts about how each
flaw behaves in the wild: how \emph{likely} exploitation is (EPSS, a statistical
score) and whether exploitation is \emph{known to be happening} (the CISA Known
Exploited Vulnerabilities catalog). Either one tightens the remediation \emph{clock};
neither changes the PAIN \emph{level}, which is driven by the CVSS vector and the
asset's requirements alone.
\end{plain}

Two governed threat inputs feed the exploitability column:
\begin{itemize}[leftmargin=1.4em]
\item \textbf{EPSS probability} $\rightarrow$ statistical likelihood. A finding at or
  above the provider-selected threshold $\theta_{\text{epss}}$ is Likely Exploitable
  (LEV) (Section~\ref{sec:remediation}).
\item \textbf{KEV membership} $\rightarrow$ confirmed exploitation. A CVE listed in
  the CISA KEV catalog is LEV regardless of its EPSS probability --- observed
  exploitation in the wild overrides any statistical estimate. KEV membership also
  independently triggers the \code{VDR-TFR-KEV} remediation dates
  (Section~\ref{sec:bod}), which run on CISA's published schedule irrespective of
  the matrix.
\end{itemize}

\noindent How it plays in --- three quick illustrations (Class~C, single-agency):
\begin{itemize}[leftmargin=1.4em]
\item \emph{KEV moves the clock (and starts a second one).} An N3 finding with EPSS
  $0.30$ (below threshold) sits in the slow column: \textbf{128 days}. It appears in
  the KEV catalog: it becomes LEV --- the deadline drops to \textbf{16 days}
  (internet-reachable) or 32 days (not) --- and the CISA-published due date applies
  in parallel under \code{VDR-TFR-KEV}.
\item \emph{EPSS moves the clock statistically.} The same N3 finding, not in the KEV
  catalog but with EPSS $0.85$ ($\ge \theta_{\text{epss}} = 0.70$), is LEV on
  likelihood alone and lands on the same 16/32-day clocks.
\item \emph{Neither present.} Sub-threshold EPSS, not in the KEV catalog, and not
  caught by the \code{FRD-LEV} unauthenticated-automation floor
  (Section~\ref{sec:remediation}) $\rightarrow$ the finding is scored purely on its
  CVSS vector and stays in the slow column at its base tier.
\end{itemize}

\section{The mathematics}
\label{sec:math}
\begin{plain}
This is the engine room --- you can skim the equations and still follow the idea.
Every flaw can hurt three things: \textbf{secrecy}, \textbf{correctness}, and
\textbf{uptime}. For each, we multiply ``how badly the flaw breaks it'' by ``how much
this system depends on it,'' then combine the three into a single damage score
between 0 and 1. Finally we translate that score into a plain word. Bigger number =
more damage. The only place human judgment enters is the three cut-points that decide
where one word becomes the next.
\end{plain}

The expected output is a PAIN level. The path is: compute an asset-specific
impact scalar $S$ from CVSS impact values and the asset's CR/IR/AR; translate
$S$ into a customer-effect word (Minimal, Narrow, Disruptive, or Debilitating);
then combine that word with the multi-agency flag to select N1--N5.

All constants below are taken \emph{verbatim} from the CVSS v3.1 specification; only
the normalization and the word thresholds (Section~\ref{sec:thresholds}) are
additions.

\begin{notation}
\begin{itemize}[leftmargin=1.3em,nosep,topsep=2pt]
\item $C, I, A$ --- how badly \emph{this flaw} breaks secrecy, correctness, and
  uptime (from the CVE's CVSS vector). None / Low / High become $0$ / $0.22$ / $0.56$.
\item $\mathrm{CR},\mathrm{IR},\mathrm{AR}$ --- how much \emph{this system} needs
  secrecy, correctness, and uptime (from its archetype). Low / Medium / High become
  $0.5$ / $1.0$ / $1.5$.
\item $\mathrm{ISC}$ --- the three combined into one ``damage to this system'' number.
\item $S$ --- that damage rescaled to a clean $0$--$1$ range.
\item $W$ --- the plain English word $S$ maps to.
\item $m$ --- does the system serve one agency ($0$) or many ($1$)? Resolved
  hierarchically from governed asset metadata.
\item $\theta_1,\theta_2,\theta_3$ --- the three cut-points where the word changes.
\end{itemize}
\end{notation}

\paragraph{Impact metric weights.}
Each of the affected vulnerability's Confidentiality, Integrity and Availability
impact metrics ($C$, $I$, $A$), read from the CVSS Base vector, takes a weight:
\[
C, I, A \in \{\,\text{None}=0,\quad \text{Low}=0.22,\quad \text{High}=0.56\,\}.
\]

\paragraph{Security Requirement weights.}
Each asset's Requirement (CR, IR, AR) takes a multiplier:
\[
\mathrm{CR},\mathrm{IR},\mathrm{AR} \in \{\,\text{Low}=0.5,\quad \text{Medium}=1.0,\quad \text{High}=1.5\,\}.
\]

\paragraph{Modified Impact Sub-Score.}
The per-dimension impacts and requirements combine as a probabilistic union --- the
share of the asset left uncompromised on each dimension is multiplied, and the
complement is the overall impact:
\begin{equation}
\mathrm{ISC} \;=\; \min\!\Big[\,1 - \big(1 - C\!\cdot\!\mathrm{CR}\big)\big(1 - I\!\cdot\!\mathrm{IR}\big)\big(1 - A\!\cdot\!\mathrm{AR}\big),\ \ 0.915\,\Big].
\label{eq:isc}
\end{equation}
The cap $0.915 = 1-(1-0.56)^3$ is the CVSS ceiling (all-High impact at Medium
requirements). We then normalize to a unit scalar:
\begin{equation}
S \;=\; \frac{\mathrm{ISC}}{0.915} \ \in\ [0,1].
\label{eq:s}
\end{equation}

\paragraph{Severity word.}
\label{sec:thresholds}
$S$ is bucketed into a FedRAMP customer-effect word by three cut points
$(\theta_1,\theta_2,\theta_3)$:
\begin{equation}
W(S) \;=\;
\begin{cases}
\text{Minimal} & S < \theta_1 \\
\text{Narrow} & \theta_1 \le S < \theta_2 \\
\text{Disruptive} & \theta_2 \le S < \theta_3 \\
\text{Debilitating} & S \ge \theta_3
\end{cases}
\qquad (\theta_1,\theta_2,\theta_3) = (0.25,\ 0.55,\ 0.80).
\label{eq:word}
\end{equation}
The cut points are the model's \emph{one} calibratable judgment. They
\rfckw{should} be documented and back-tested against analyst-rated findings, and an
implementation \rfckw{must} treat them as governed configuration rather than an
in-band, ad-hoc setting.

\paragraph{Scope and the N-level.}
Let $m \in \{0,1\}$ be the asset's multi-agency flag, resolved from governed
asset metadata using the most-specific applicable value. Scope is purely
hierarchical --- an asset is multi-agency only if the asset itself, a containing
boundary, or the provider default is tagged multi-agency. The N-level follows
from the word and scope:
\begin{equation}
\text{PAIN} =
\begin{cases}
\text{N1} & W=\text{Minimal} \\
\text{N2} & W=\text{Narrow} \\
\text{N3} & W=\text{Disruptive},\ m=0 \\
\text{N4} & W=\text{Disruptive},\ m=1 \ \ \text{or}\ \ W=\text{Debilitating},\ m=0 \\
\text{N5} & W=\text{Debilitating},\ m=1
\end{cases}
\label{eq:pain}
\end{equation}

\section{Fail-safe behavior}
Missing metadata \rfckw{must not} \emph{lower} PAIN. If a system has not been
classified yet, the method assumes the worst rather than the best: an asset that
carries no classification \rfckw{should} resolve to a deliberately conservative
default (e.g.\ CR/IR/AR all High), which scores loudly and surfaces the asset for
classification rather than hiding it. ``Unknown'' is treated as ``serious until
proven otherwise'' --- silence never makes a problem look smaller than it is.

\section{Remediation: the VDR-TFR-PVR matrix}
\label{sec:remediation}
\begin{plain}
The N-level says \emph{how bad}; this step says \emph{how fast}. Two real-world facts
tighten the clock: is the flaw actually being exploited (or very likely to be), and
is the system reachable from the internet? More severe + more exploitable + more
exposed = less time to fix. A lookup table turns those into a concrete deadline,
stricter for higher-assurance providers.
\end{plain}
The remediation deadline is selected by three values: the provider Certification
Class, the PAIN level, and an exploitability \emph{column} derived from the two VER
axes:
\[
\text{deadline} = M\big[\,\text{Class}\,\big]\big[\,\text{PAIN}\,\big]\big[\,\text{column}\,\big],
\qquad
\text{column} \in \{\,\text{LEV+IRV},\ \text{LEV+NIRV},\ \text{NLEV}\,\}.
\]
where, for a provider-selected EPSS threshold $\theta_{\text{epss}}$ (this memo uses
$0.70$):
\begin{align*}
\text{LEV} \;&=\; \big(\text{EPSS} \ge \theta_{\text{epss}}\big)\ \lor\ \big(\text{CVE} \in \text{KEV}\big)\ \lor\ \big(\text{IRV}_{\text{direct}} \land \text{unauthenticated automation}\big),\\
\text{IRV} \;&=\; \text{the \emph{finding} is internet-reachable (evaluated per vulnerability, not per asset).}
\end{align*}
The third LEV disjunct is not a provider option: it implements the \code{FRD-LEV}
floor note that ``any vulnerability that an automated unauthenticated system can
exploit over the internet is a likely exploitable vulnerability.'' A deterministic
proxy suffices: the finding is IRV via \emph{direct} exposure and its CVSS vector
permits unauthenticated automation (\code{AV:N/PR:N/UI:N}).

\begin{definitionbox}
Internet reachability is a property of the \emph{(vulnerability, asset) pair}, not
of the asset alone. Per \code{FRD-IRV}, an internet-reachable vulnerability is one
that ``might be exploited or otherwise triggered by a payload originating from a
source on the public internet'' --- explicitly including resources with \emph{no
direct internet route} that process payloads received indirectly;
\code{VER-EVA-EIR}'s canonical example is SQL injection reaching a database on a
private network through the application tier. The determination is therefore made
per finding:
\[
\mathrm{IRV}(v,a) \;=\; \big[\,\text{direct exposure} \ \lor\ \text{transitive payload exposure}\,\big]
\ \land\ \text{per-vulnerability filters},
\]
where the filters are execution evidence, trigger-surface match, and trigger
mechanism (content-triggered vs.\ peer-bound), and \emph{every term defaults to
reachable} absent deterministic evidence. Two corrections to common intuitions:
(a) \code{FRD-IRV} imposes no unauthenticated requirement --- ``unauthenticated
automated exploitation'' belongs to the \code{FRD-LEV} exploitability floor, not
to reachability; (b) edge authentication (an identity-aware proxy and a client VPN
alike) excludes a backend from IRV only when it functions as a
\emph{remote-access boundary}: the boundary rejects unauthenticated traffic before
the vulnerable code processes it, the authorized principal set is restricted,
enumerable, and organization-controlled, and the backend validates the boundary's
signed identity assertion. Edge authentication functioning as customer
authentication for the service itself (its general internet user base) is an
exploitability signal, not an IRV exclusion. The full
discovery method is specified in the companion paper \emph{An Exclusion-Gated
Method for Determining Internet Reachability of Vulnerabilities under the FedRAMP
VDR/VER Rules}.
\end{definitionbox}

Restricting the threat inputs to EPSS and the KEV catalog --- and to nothing else ---
is itself a design opinion (see Appendix~\ref{app:opinions}). The day-valued tables
are given in Appendix~\ref{app:matrix}. PAIN-1 carries no FedRAMP deadline.

\section{Worked examples}
\begin{plain}
The same formula, run by hand on realistic cases. Watch the \emph{same}
vulnerability land at very different N-levels depending only on which system it sits
on --- that is the entire point of the method.
\end{plain}
We use $C,I,A$ for impacts and CR/IR/AR for requirements; all arithmetic follows
Eqs.~\eqref{eq:isc}--\eqref{eq:pain}.

For reference, the severity word used in each example is assigned by:
\[
W(S) \;=\;
\begin{cases}
\text{Minimal} & S < \theta_1 \\
\text{Narrow} & \theta_1 \le S < \theta_2 \\
\text{Disruptive} & \theta_2 \le S < \theta_3 \\
\text{Debilitating} & S \ge \theta_3
\end{cases}
\qquad
(\theta_1,\theta_2,\theta_3) = (0.25,\ 0.55,\ 0.80).
\]

\paragraph{Example 1 --- RCE on a crown-jewel datastore, multi-agency.}
$C\!=\!I\!=\!A\!=\!0.56$ (High), CR=IR=AR=$1.5$ (High), $m=1$.
\begin{align*}
\mathrm{ISC} &= \min\!\big[\,1-(1-0.84)^3,\ 0.915\,\big] = \min[\,0.9959,\ 0.915\,] = 0.915, \\
S &= 0.915/0.915 = 1.000 \;\Rightarrow\; W=\text{Debilitating}, \\
m=1 \;\Rightarrow\; \boxed{\text{N5}}.
\end{align*}

\paragraph{Example 2 --- the same CVE on a sandbox, single-agency.}
$C\!=\!I\!=\!A\!=\!0.56$, CR=IR=AR=$0.5$ (Low), $m=0$.
\begin{align*}
\mathrm{ISC} &= 1-(1-0.28)^3 = 1-0.72^3 = 0.6268, \\
S &= 0.6268/0.915 = 0.685 \;\Rightarrow\; \text{Disruptive} \;\Rightarrow\; \boxed{\text{N3}}.
\end{align*}
Identical vulnerability, opposite ends of the matrix --- the asset half of the
score (CR/IR/AR) is doing the work.

\paragraph{Example 3 --- availability-only DoS on a message backbone.}
A CVE with $C\!=\!0,\ I\!=\!0,\ A\!=\!0.56$ (e.g.\ a parser DoS), on an asset with
CR=IR=AR=$1.5$, $m=0$. EPSS $=0.76$. The vulnerable parser sits on the
inter-broker replication surface, which receives no internet-derived data, so the
finding is not internet-reachable (taint-excluded under the companion method ---
had the flaw been in \emph{message-payload} parsing it would be transitively IRV).
\begin{align*}
\mathrm{ISC} &= 1-(1-0)(1-0)(1-0.84) = 1-0.16 = 0.84, \\
S &= 0.84/0.915 = 0.918 \;\Rightarrow\; W=\text{Debilitating},\ m=0 \;\Rightarrow\; \boxed{\text{N4}}.
\end{align*}
The single dimension $A\!\cdot\!\mathrm{AR}=0.84$ alone clears the Debilitating bar.
Remediation: LEV (EPSS $\ge 0.70$) $\land$ NIRV; at Class C this is
$M[\text{C}][\text{N4}][\text{LEV+NIRV}] = 8$ days. Note this finding may carry a
\emph{Low} vendor severity --- PAIN is decoupled from the qualitative label.

\paragraph{Example 4 --- information disclosure on a metadata-only service.}
$C\!=\!0.56,\ I\!=\!0,\ A\!=\!0$, on an asset with CR=$0.5$ (Low), IR=AR=$1.5$.
\begin{align*}
\mathrm{ISC} &= 1-(1-0.28)(1)(1) = 0.28, \\
S &= 0.28/0.915 = 0.306 \;\Rightarrow\; W=\text{Narrow} \;\Rightarrow\; \boxed{\text{N2}}.
\end{align*}
A confidentiality leak on a service that holds only operational metadata is
reconnaissance, not a debilitating breach --- captured by the low CR.

\section{Mitigation and the PAIN Ladder}
\label{sec:ladder}
\begin{plain}
FedRAMP does not demand that every flaw be \emph{fixed} inside its deadline --- it
demands that the potential damage come \emph{down the ladder} on schedule. There are
two ways to climb down: make the flaw hurt less if it fires (its \emph{row} drops to
a lower N-level), or make it harder to reach or exploit (its \emph{column} moves to a
slower clock). Every step down must be backed by evidence somebody signed, every step
is recorded, and no step may claim the damage is exactly zero --- ``zero'' is a
different kind of claim with its own paperwork (VEX).
\end{plain}

FedRAMP's response model is a ladder, not a binary. \code{VER-RPT-VDT} requires
reporting the ``time and Potential Agency Impact N-rating of each completed and
evaluated reduction in Potential Agency Impact N-rating'' and the ``estimated time
and target Potential Agency Impact N-rating of next reduction in Potential Agency
Impact N-rating'' --- the rule \emph{expects} PAIN to be re-estimated downward as
mitigations land --- and the \code{VDR-TFR-PVR} timeframes are satisfied by partial
mitigation to a lower N-rating (Appendix~\ref{app:matrix}). A deterministic
classifier therefore needs a deterministic \emph{lever}: a governed way to recompute
PAIN when a mitigation changes the facts.

\paragraph{The lever is the other half of CVSS Environmental.} Section~\ref{sec:math}
uses one half of the Environmental group, the Security Requirements (CR/IR/AR). The
other half is the \emph{Modified Base} metrics, which the CVSS specification provides
precisely for capturing mitigations, configurations, and compensating controls
present in a specific environment. The two halves split cleanly along the model's
own seam:
\begin{itemize}[leftmargin=1.4em,nosep,topsep=3pt]
\item \textbf{Asset side --- static.} CR/IR/AR and the multi-agency flag describe
  what the asset \emph{is}. They never move to make a deadline.
\item \textbf{Vulnerability side --- dynamic.} The Modified impact metrics
  (\code{MC}/\code{MI}/\code{MA}) describe what the flaw can still do \emph{here},
  given the controls actually deployed. They are evidence-gated and move as
  mitigations land.
\end{itemize}
Recomputation runs through the same arithmetic --- substitute the modified values
for $C,I,A$ in Eq.~\eqref{eq:isc} --- and the word, the N-level, and the deadline
follow exactly as before. No new formula, no side channel.

\paragraph{Concrete row-movers.} These are ordinary hardening actions, not magic
wands; each has a defensible causal story for the dimension it lowers:
\begin{itemize}[leftmargin=1.4em,nosep,topsep=3pt]
\item Drop the vulnerable process to a non-root user with a read-only filesystem ---
  the effective Integrity (and often Confidentiality) of an RCE falls.
\item Scope the application's database credential to a single schema --- a SQL
  injection's C/I falls: the payload still arrives, but it reaches less.
\item Disable the vulnerable feature --- the impact of a code path that cannot be
  invoked falls. (Removing the vulnerable code \emph{outright} goes further and
  graduates to a VEX disposition; see below.)
\item Confine egress --- default-deny outbound rules break exploit chains that
  need a second stage.
\item Rate-limit the affected endpoint --- Availability impact falls for
  volume-dependent flaws.
\end{itemize}

\paragraph{Worked example: Log4Shell egress-deny.} In December 2021, most
organizations' first effective Log4Shell mitigation was not a WAF. One of the most
widely deployed was default-deny egress --- a NetworkPolicy or security-group change
blocking outbound LDAP/RMI from application subnets. Log4Shell's RCE requires a
second-stage fetch of the malicious class from the attacker's LDAP server, so
blocking egress breaks the chain; but DNS-based exfiltration of environment
variables (\code{\$\{env:AWS\_SECRET\_ACCESS\_KEY\}}) remained observed in the wild
--- which is exactly why the Modified Confidentiality drops to Low and never to
None. The arithmetic runs through Eqs.~\eqref{eq:isc}--\eqref{eq:pain} unchanged.
Base, with $C\!=\!I\!=\!A\!=\!0.56$ (High) on a CR=IR=AR=$1.5$ (High) multi-agency
asset:
\begin{align*}
\mathrm{ISC} &= \min\!\big[\,1-(1-0.84)^3,\ 0.915\,\big] = 0.915,\quad
S = 1.000 \;\Rightarrow\; \text{Debilitating},\ m=1 \;\Rightarrow\; \text{N5},
\end{align*}
on the Class~C LEV+IRV clock of \textbf{2 days} (Log4Shell is LEV twice over:
KEV-listed, and caught by the \code{FRD-LEV} floor). With the evidenced egress-deny,
$\mathrm{MC}=\mathrm{MI}=\mathrm{MA}=0.22$ (Low):
\begin{align*}
\mathrm{ISC} &= 1-(1-0.33)^3 = 0.699,\quad
S = 0.699/0.915 = 0.764 \;\Rightarrow\; \text{Disruptive},\ m=1 \;\Rightarrow\;
\boxed{\text{N4}}.
\end{align*}
The N5 timeframe is satisfied by the reduction, the reduction is reported under
\code{VER-RPT-VDT}, and work proceeds on the N4 clock. The evidence is the
NetworkPolicy object snapshot and the data-flow graph showing the egress edge
removed (the companion method's telemetry). One clock is untouched: Log4Shell is a
KEV, so the \code{VDR-TFR-KEV} remediation date runs independently of any
reduction.

\paragraph{Guardrails.} The lever only ever lowers apparent risk, so it carries the
same governance as the disposition overlay of Section~\ref{sec:vex}:
\begin{enumerate}[leftmargin=1.7em,nosep,topsep=3pt]
\item A modification \rfckw{may} only \emph{lower} a dimension on cited, signed
  evidence, and \rfckw{must} be re-evaluated whenever the control it relies on
  changes. The evidence bar scales with the PAIN being reduced.
\item A modification \rfckw{may} lower High to Low but \rfckw{must not} lower any
  dimension to None. Claiming \emph{zero} impact is an applicability claim, and it
  travels as a VEX disposition (\code{not\_affected}), not as a metric modification.
\item The audit trail records base $\rightarrow$ modified per dimension with the
  evidence references, mirroring the retained-PAIN pattern: the pre-mitigation
  rating stays on record and the reduction is attributable and reversible.
\end{enumerate}

\paragraph{The rest of the menu, from the same incidents.} The guardrails are not
hypothetical --- each has already played out in public:
\begin{itemize}[leftmargin=1.4em,nosep,topsep=3pt]
\item \textbf{Code removal (Log4Shell).} Apache's own interim recommendation
  removes the vulnerable class from the jar entirely:\\[2pt]
  {\footnotesize\code{zip -q -d log4j-core-*.jar org/apache/logging/log4j/core/lookup/JndiLookup.class}}\\[2pt]
  In this model that claim travels as a VEX disposition
  (\code{vulnerable\_code\_not\_present}) evidenced by the modified jar hash --- not
  as a Modified-impact claim (guardrail~2).
\item \textbf{Config kill-switch, and the re-evaluation lesson (Log4Shell).} The
  JVM flag \code{-Dlog4j2.formatMsgNoLookups=true} was initially treated as a
  complete mitigation --- until CVE-2021-45046 showed non-default configurations (Thread
  Context Map patterns) where lookups still fired, and the flag was downgraded to
  insufficient. That is guardrail~1 playing out in the real world: mitigations carry
  timestamps and are re-evaluated when the facts change; a one-time silent
  suppression would have outlived its justification.
\item \textbf{regreSSHion (CVE-2024-6387).} The published mitigation is one line of
  \code{sshd\_config}: \code{LoginGraceTime 0}, closing the pre-authentication RCE
  race at the cost of a connection-exhaustion DoS risk --- a mitigation that
  \emph{trades} impact dimensions (Modified C/I fall, Modified A is retained or
  worsens), and one no WAF could ever address.
\item \textbf{Spring4Shell (CVE-2022-22965).} Spring's interim workaround was a
  \code{@ControllerAdvice} restricting the data binder's \code{disallowedFields}
  (\code{"class.*"} and friends) --- an application-configuration mitigation shipped
  while framework upgrades were scheduled.
\end{itemize}

\paragraph{Column moves: the other axis.} Reachability and exploitability
mitigations do not touch the impact metrics at all --- they move the finding's
\emph{column}. Network segmentation that verifiably cuts the payload path (per the
companion Exclusion-Gated method's data-flow evidence) flips IRV to NIRV; a
qualifying remote-access boundary (the companion method's Gate~3) defeats the
\code{FRD-LEV} unauthenticated-automation floor. \code{VDR-TFR-PVR} is satisfied by landing on
\emph{any} longer clock, whether the step down was a row or a column.

\paragraph{Where the ladder ends.} The descent is bounded, not endless. It
terminates in one of four states: \emph{remediation}; \emph{full mitigation}
(\code{FRD-FMV}: the vulnerability ``will still exist (with negligible risk) until
it has been remediated,'' the remaining remediation handled in routine operations
per \code{VDR-TFR-RMN}, KEVs excepted); \emph{PAIN-1}, which carries no timeframe
(Appendix~\ref{app:matrix}) and likewise returns the finding to routine operations;
or \emph{formal acceptance} --- \code{VER-TFR-MAV} requires categorizing any
vulnerability that is not or will not be fully mitigated or remediated within 192
days of evaluation as an accepted vulnerability, tracked on the accepted track of
Section~\ref{sec:vex}. It is no coincidence that 192 days is also the largest cell
in the timeframe matrix: the grid and the acceptance boundary meet by construction.
Nor does each rung demand a fresh mitigation. Each step lands on a strictly longer
clock (at Class~C, LEV$+$IRV: $2 \rightarrow 4 \rightarrow 16 \rightarrow 48$
days), so the common pattern is a single mitigation that buys schedule, followed by
remediation on the longer clock --- or, where remediation is genuinely impossible,
acceptance at the 192-day boundary rather than an indefinite treadmill.

\paragraph{WAF realism.} A WAF virtual patch is one item on this menu, not the
presumed answer. \code{VER-EVA-EIR} calls payload interception ``the simplest way''
to prevent exploitation --- not the expected way. Narrow, CVE-specific virtual
patches are often practical (blocking Log4Shell's JNDI lookup strings); generic
rules --- say, parameterized SQL-injection signatures across a real API surface ---
carry false-positive risk to legitimate customer traffic, and the claimed mitigation
\rfckw{must} be re-validated whenever the rule set changes. The incident record
above bears this out: under real deadline pressure, the mitigations that actually
shipped were configuration flags, code removal, privilege and egress confinement,
and binding restrictions --- surgical, testable, and zero risk to customer traffic
--- while the WAF rule was the fallback for the workloads that could not be touched
(appliances, vendor binaries). That is precisely its position in this method. The
companion Exclusion-Gated paper develops the corresponding posture recommendation:
keep the finding IRV and attest the perimeter mitigation via signed VEX rather than
reclassifying reachability on WAF evidence.

\section{Reference architectures and sample scans}
\subsection{Single-agency, Class C}
Figure~\ref{fig:arch} shows a representative deployment with each component tagged
by archetype. Table~\ref{tab:scan} shows the corresponding per-finding output for a
fictional Class~C, single-agency offering. The PAIN column is computed by the method
above; the Classical Severity column is the scanner's qualitative rating, shown to
make the decoupling explicit.

\begin{figure}[H]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  font=\sffamily\small,
  node distance=10mm and 14mm,
  box/.style={draw=rule,rounded corners,fill=white,minimum height=9mm,minimum width=24mm,align=center,thick},
  edge/.style={-{Stealth[length=2mm]},draw=infoblue,thick},
  tag/.style={font=\ttfamily\scriptsize,fill=tagbg,inner sep=1.5pt,rounded corners}
]
\node[box] (net) {Internet};
\node[box,below=of net] (ing) {Ingress / LB};
\node[box,below=of ing] (web) {Web frontend};
\node[box,right=26mm of web] (auth) {Auth + Redis\\(sessions)};
\node[box,below=of web] (api) {API service};
\node[box,below=of api] (kafka) {Kafka};
\node[box,right=26mm of kafka] (pg) {Postgres};
\node[box,left=22mm of api] (dns) {CoreDNS};
\node[box,left=22mm of kafka] (cicd) {CI/CD runner};
\node[box,right=26mm of auth] (siem) {SIEM / EDR};

\draw[edge] (net) -- (ing);
\draw[edge] (ing) -- (web);
\draw[edge] (web) -- (api);
\draw[edge] (web) -- (auth);
\draw[edge] (api) -- (kafka);
\draw[edge] (api) -- (pg);
\draw[edge] (kafka) -- (pg);
\draw[edge,dashed] (cicd) -- (api);
\draw[edge,dashed] (dns) -- (api);

\node[tag,below=0.5mm of net.south] {n/a};
\node[tag,below=0.5mm of ing.south] {public-edge};
\node[tag,below=0.5mm of web.south] {app-tier};
\node[tag,below=0.5mm of auth.south] {identity-secrets};
\node[tag,below=0.5mm of api.south] {app-tier};
\node[tag,below=0.5mm of kafka.south] {data-backbone};
\node[tag,below=0.5mm of pg.south] {data-sensitive};
\node[tag,below=0.5mm of dns.south] {platform-foundation};
\node[tag,below=0.5mm of cicd.south] {cicd-pipeline};
\node[tag,below=0.5mm of siem.south] {security-tooling};
\end{tikzpicture}%
}
\caption{Reference architecture; each node carries its \code{asset-archetype} tag.}
\label{fig:arch}
\end{figure}

\begin{table}[H]
\centering
\renewcommand{\arraystretch}{1.3}
\setlength{\tabcolsep}{4pt}
\footnotesize
\begin{tabular}{l l c c c c r}
\toprule
\textbf{CVE} & \textbf{Resource (archetype)} & \textbf{Classical} & \textbf{EPSS} & \textbf{IRV} & \textbf{PAIN} & \textbf{Remediation} \\
\midrule
CVE-2026-xxx1 & ingress \tagn{(public-edge)}        & \cellcolor{crit!25}CRITICAL & 0.97 & yes & \cellcolor{high!35}N4 & 4 days \\
CVE-2026-xxx2 & payments-db \tagn{(data-sensitive)} & \cellcolor{crit!25}CRITICAL & 0.94 & yes\,(t) & \cellcolor{high!35}N4 & 4 days \\
CVE-2026-xxx3 & redis \tagn{(identity-secrets)}     & \cellcolor{high!25}HIGH     & 0.50 & no  & \cellcolor{high!35}N4 & 64 days \\
CVE-2026-xxx4 & kafka \tagn{(data-backbone)}        & \cellcolor{low!30}LOW       & 0.76 & no  & \cellcolor{high!35}N4 & 8 days \\
CVE-2026-xxx5 & web \tagn{(app-tier)}               & \cellcolor{med!35}MEDIUM    & 0.20 & yes & \cellcolor{med!35}N3 & 128 days \\
CVE-2026-xxx6 & coredns \tagn{(platform-foundation)}& \cellcolor{med!35}MEDIUM    & 0.08 & no  & \cellcolor{low!30}N2 & 192 days \\
\bottomrule
\end{tabular}
\caption{Sample VDR scan output (fictional data, placeholder CVE IDs; Class~C, single-agency). PAIN is
computed from the archetype's CR/IR/AR and the CVE impact vector; the deadline is
$M[\text{C}][\text{PAIN}][\text{column}]$. The IRV column is evaluated \emph{per
finding} via the companion Exclusion-Gated method; (t) marks transitive payload
exposure. CVE-2026-xxx2 is FedRAMP's canonical case: a payload-carried
(SQL-injection-class) finding on a datastore that is not internet-accessible is
still IRV, tightening the deadline to 4 days. CVE-2026-xxx4 --- a \emph{Low}
vendor severity that still yields N4 (availability-High DoS on a
High-availability-requirement asset; Example~3) --- stays NIRV because its
vulnerable parser sits on the inter-broker replication surface, which receives no
internet-derived data.}
\label{tab:scan}
\end{table}

\subsection{Multi-agency, Class D}
Figure~\ref{fig:archd} is a shared, multi-tenant platform serving \emph{several}
agencies at the highest assurance (Class~D). Its shared components carry more than
one agency's data, so they are tagged multi-agency (here via the cluster default,
with single-agency tenant workloads tagged as the exceptions). Table~\ref{tab:scand}
shows the result: the same families of flaws now remediate in \emph{hours}.

\begin{figure}[H]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  font=\sffamily\small,
  node distance=10mm and 14mm,
  box/.style={draw=rule,rounded corners,fill=white,minimum height=9mm,minimum width=24mm,align=center,thick},
  edge/.style={-{Stealth[length=2mm]},draw=infoblue,thick},
  tag/.style={font=\ttfamily\scriptsize,fill=tagbg,inner sep=1.5pt,rounded corners}
]
\node[box] (net) {Internet};
\node[box,below=of net] (gw) {API gateway};
\node[box,below=of gw] (app) {Tenant pods};
\node[box,right=26mm of app] (idp) {Shared IdP\\+ KMS};
\node[box,below=of app] (bus) {Shared event bus};
\node[box,right=26mm of bus] (db) {Agency datastores};
\node[box,left=22mm of app] (dns) {Shared DNS};
\node[box,left=22mm of bus] (argo) {GitOps / ArgoCD};
\node[box,right=26mm of idp] (audit) {Audit log store};

\draw[edge] (net) -- (gw);
\draw[edge] (gw) -- (app);
\draw[edge] (app) -- (idp);
\draw[edge] (app) -- (bus);
\draw[edge] (bus) -- (db);
\draw[edge,dashed] (dns) -- (app);
\draw[edge,dashed] (argo) -- (app);

\node[tag,below=0.5mm of net.south] {n/a};
\node[tag,below=0.5mm of gw.south] {public-edge};
\node[tag,below=0.5mm of app.south] {app-tier};
\node[tag,below=0.5mm of idp.south] {identity-secrets};
\node[tag,below=0.5mm of bus.south] {data-backbone};
\node[tag,below=0.5mm of db.south] {data-sensitive};
\node[tag,below=0.5mm of dns.south] {platform-foundation};
\node[tag,below=0.5mm of argo.south] {config-actuation};
\node[tag,below=0.5mm of audit.south] {security-tooling};
\end{tikzpicture}%
}
\caption{Multi-agency reference architecture; shared components serve several
agencies and are tagged multi-agency.}
\label{fig:archd}
\end{figure}

\begin{table}[H]
\centering
\renewcommand{\arraystretch}{1.3}
\setlength{\tabcolsep}{4pt}
\footnotesize
\begin{tabular}{l l c c c c r}
\toprule
\textbf{CVE} & \textbf{Resource (archetype)} & \textbf{Classical} & \textbf{EPSS} & \textbf{IRV} & \textbf{PAIN} & \textbf{Remediation} \\
\midrule
CVE-2026-yyy1 & gateway \tagn{(public-edge)}         & \cellcolor{crit!25}CRITICAL & 0.90 & yes & \cellcolor{crit!30}N5 & 12 hours \\
CVE-2026-yyy2 & shared-idp \tagn{(identity-secrets)} & \cellcolor{crit!25}CRITICAL & 0.95 & no  & \cellcolor{crit!30}N5 & 1 day \\
CVE-2026-yyy3 & event-bus \tagn{(data-backbone)}     & \cellcolor{low!30}LOW       & 0.80 & yes\,(t) & \cellcolor{crit!30}N5 & 12 hours \\
CVE-2026-yyy4 & agency-db \tagn{(data-sensitive)}    & \cellcolor{high!25}HIGH     & 0.40 & no  & \cellcolor{crit!30}N5 & 8 days \\
CVE-2026-yyy5 & tenant-pod \tagn{(app-tier)}         & \cellcolor{med!35}MEDIUM    & 0.10 & no  & \cellcolor{med!35}N3 & 64 days \\
CVE-2026-yyy6 & shared-dns \tagn{(platform-foundation)} & \cellcolor{med!35}MEDIUM & 0.05 & no  & \cellcolor{low!30}N2 & 192 days \\
\bottomrule
\end{tabular}
\caption{Sample VDR scan output (fictional data, placeholder CVE IDs; Class~D, multi-agency). Multi-agency
scope escalates Disruptive/Debilitating findings, and Class~D carries the tightest
deadlines --- an internet-facing exploitable flaw on the shared edge remediates in
\textbf{12 hours} (CVE-2026-yyy1). CVE-2026-yyy3 shows a \emph{Low} vendor severity
at N5 also remediating in \textbf{12 hours}: a payload-carried DoS in
message-payload parsing on the shared backbone is transitively internet-reachable
(t) through the tenant pods, even though the bus itself is not internet-accessible.
A confidentiality leak on shared DNS (CVE-2026-yyy6) stays N2 and NIRV ---
``Narrow'' does not escalate with scope.}
\label{tab:scand}
\end{table}

\section{Relationship to CISA BOD 26-04}
\label{sec:bod}
The Rev5 VDR/VER rule is FedRAMP's implementation of CISA Binding Operational
Directive~26-04.\footnote{CISA, \emph{Binding Operational Directive 26-04}, 2026-06-10.
The VDR rule lists it as the mandating authority, with optional adoption from
2026-07-04, obtain/maintain from 2026-12-07, and grace ending 2027-03-07.} FedRAMP
adopts the directive's \emph{philosophy} --- move remediation prioritization off raw
CVSS base scores and onto exploitability and mission impact --- but substitutes its
own instruments for the directive's pillars:

\begin{table}[H]
\centering
\renewcommand{\arraystretch}{1.25}
\footnotesize
\begin{tabular}{>{\raggedright\arraybackslash}p{0.30\textwidth} >{\raggedright\arraybackslash}p{0.30\textwidth} >{\raggedright\arraybackslash}p{0.30\textwidth}}
\toprule
\textbf{BOD 26-04 pillar} & \textbf{FedRAMP VDR/VER instrument} & \textbf{Status} \\
\midrule
SSVC drives prioritization & \textbf{PAIN} (Potential Agency Impact, N1--N5) & SSVC not named; PAIN replaces it \\
VEX communicates exploitability & \textbf{LEV} (exploitation) + disposition (\S\ref{sec:vex}) & VEX not named; role split across LEV and VER disposition \\
CSAF machine-readable advisories & VER JSON reporting (\S\ref{sec:vex}) & format unspecified; CycloneDX VEX proposed (companion) \\
KEV catalog due dates & \texttt{VDR-TFR-KEV} (defers to CISA dates) & retained intact; also consumed as a LEV input \\
(none) & \textbf{IRV/NIRV} and Certification \textbf{Class} & added as timeframe axes \\
\bottomrule
\end{tabular}
\caption{FedRAMP keeps BOD~26-04's intent but swaps SSVC$\,\rightarrow\,$PAIN and
VEX$\,\rightarrow\,$LEV${+}$disposition, retains KEV, and adds reachability and Class.}
\end{table}

The first row is the deviation that shapes this memo. Because FedRAMP chose
\emph{PAIN} rather than SSVC, an SSVC-based prioritizer does not by itself satisfy the
rule --- and FedRAMP publishes PAIN's \emph{output} buckets and the timeframe table
but not the \emph{classifier} that produces an N-level. That gap is exactly what
Section~\ref{sec:math} fills, which is why this method is expressed in CVSS
Environmental terms (the missing input was PAIN, not an exploitability score) rather
than in SSVC. The directive's other instruments retain narrower roles: this memo
consumes the \emph{KEV catalog} directly as one input to LEV
(Section~\ref{sec:remediation}) --- arguably a cleaner story than substituting for
it, since BOD~26-04's own exploitation instrument is kept intact rather than
replaced by a derived signal --- while VEX carries the post-detection
\emph{disposition} of each finding (Section~\ref{sec:vex}). Neither SSVC nor VEX is
mandated by the rule by name.

\section{Assigning CR/IR/AR: archetypes (an example, not a standard)}
\label{sec:archetypes}
\begin{plain}
Rather than hand-rate every server, we sort systems into a short list of
\emph{archetypes} --- a database, a login service, a cache, a message bus, and so on
--- and each archetype comes with its criticality pre-filled. Tag a system with its
archetype and the scoring is automatic. Our list is only an \emph{example}: every
provider should tailor its own to the data it actually holds.
\end{plain}
The model in Section~\ref{sec:math} is only as good as the CR/IR/AR assignment that
feeds it. Without an archetype approach, a provider would have to tag each asset
with CR/IR/AR scores individually. While certainly justifiable and valid, this
requires evaluating each asset in isolation. In doing so, we lose a systematic ``chain
of thought'' --- for instance, why is a specific host scored CR:L, IR:H, AR:L?

By assigning requirements to archetypes instead, the underlying rationale remains
readily understandable and easily auditable. Tagging an asset with its archetype
inherits that clear chain of thought. We propose the archetype taxonomy herein as
a valid option, while recognizing that not every asset will fit cleanly. CSPs
\rfckw{may} define custom archetypes or adjust the scoring of existing ones at
their discretion to map their estate accurately. The goal is to identify the most
critical assets as defined by FedRAMP and deprioritize less critical ones.


\begin{quote}\itshape
The archetype catalog in Appendix~\ref{app:catalog} is illustrative. A conforming
implementation \rfckw{must} derive its CR/IR/AR assignment from its own system
categorization (FIPS-199 impact levels, data inventory, authorization boundary). It
\rfckw{may} adopt the example catalog as a starting point, but each CSP
\rfckw{should} own and justify its mapping. Two providers with different data
inventories can legitimately assign the same software different requirements.
\end{quote}

The classification rule we recommend is \emph{control-plane lens first}: if an
asset can deploy, orchestrate, hold cross-estate credentials, or actuate
configuration, classify it by that control function regardless of the data it
stores; otherwise classify by the data it holds. The same physical software lands
in different archetypes by role --- an in-memory store used as a cache is
\code{app-tier}, the same software used as a session/token store is
\code{identity-secrets}, and used as a job broker is \code{data-backbone}.

For Kubernetes estates, the multi-agency flag can be resolved most-specific-first:
workload label, then namespace label, then cluster default. Other deployment
models should apply the same rule to their own asset, boundary, and provider-default
metadata.

\section{Applicability and the disposition via Vulnerability Exploitability eXchange (VEX)}
\label{sec:vex}
\begin{plain}
Scoring says how bad a finding is and how fast to fix it. But a scanner's hit is not
yet a confirmed problem: the flagged code may not be present, may never run, may need
a configuration nobody set, or may already be neutralized by a control. Deciding which
is the \emph{disposition} step --- and it can only happen \emph{after} a finding is
detected, because you cannot pre-judge factors specific to a vulnerability you have not
yet seen. This section places that step in the method and names the kind of artifact
that records it: VEX.
\end{plain}

The PAIN derivation in Sections~\ref{sec:math} and~\ref{sec:remediation} is
deterministic: given the asset's CR/IR/AR, the VER axes, and the configured
thresholds, the N-level and deadline follow by formula. Disposition answers a
different question: whether the scanner finding is actually exploitable in this
running environment. That requires evidence about the specific vulnerability and
deployment --- for example, whether the affected code is present, reachable,
configured, or already neutralized by a compensating control. The analysis may be
automated, but it happens after detection because those facts are finding-specific.

\paragraph{Where FedRAMP requires it.} VER requires three things relevant here:
evaluate whether a scanner finding is real, record the per-finding disposition, and
track vulnerabilities that are accepted rather than remediated. The FedRAMP clauses
behind those duties are \code{VER-EVA-EFP}, \code{VER-EVA-AIA},
\code{VER-RPT-VDT}, and \code{VER-TFR-MAV}/\code{VER-RPT-AVI}. VER makes the
disposition step mandatory, but it does not prescribe a carrier format. The
\emph{Vulnerability Exploitability eXchange} (VEX) is a standardized,
machine-readable, signable artifact for carrying those assertions.

\paragraph{Two outcomes.} A disposition resolves a finding onto one of two tracks:
\begin{itemize}[leftmargin=1.5em,nosep,topsep=3pt]
\item \textbf{Not truly affected} --- suppressed with a machine-readable
  \emph{justification} (component or vulnerable code not present, code not reachable,
  exploitation requires a configuration that is not set, or a compensating control is
  already in place).
\item \textbf{Affected but remediation will be late} --- under VER's tight clocks a
  provider \rfckw{may} be unable to patch before the deadline; the finding is recorded
  as \emph{accepted}, with an attested mitigation and a planned response.
\end{itemize}

\paragraph{A disposition overlay, not a re-score.} Disposition does not change the
arithmetic. Every detected finding is scored as in Section~\ref{sec:math}; a VEX
assertion then moves the finding from the active remediation queue into a suppressed or
accepted set \emph{while its computed PAIN is retained}. Keeping the would-be PAIN on
record makes every suppression auditable and reversible --- the safeguard the Security
Considerations (Section~\ref{sec:seccon}) call for: a downgrade by disposition is
visible, attributable, and can be revisited if the assertion is later contradicted.

\paragraph{A distinct axis.} Applicability (VEX) is orthogonal to
exploitation-likelihood (LEV, Section~\ref{sec:remediation}). LEV asks \emph{how likely
is this vulnerability to be exploited in the wild}; VEX asks \emph{is this product, as
deployed, affected at all}. A finding can be likely-exploitable in general yet
\code{not\_affected} here because the vulnerable path is unreachable. The two are
evaluated independently and \rfckw{must not} be collapsed into one.

\paragraph{Guardrails.} Because a disposition only ever \emph{lowers} apparent risk ---
and the more so once the determination is automated --- the governing controls matter as
much as the analysis. Absent sufficient evidence a finding \rfckw{must} remain affected,
honoring \code{VER-EVA-AIA}; and the evidence required to suppress \rfckw{should} scale
with the retained PAIN, so a KEV-class, internet-reachable, high-$N$ finding demands far
stronger corroboration --- and human review --- than a low-$N$ internal one. Every
disposition \rfckw{should} cite the evidence it rests on and be signed, so it is
reproducible and attributable; and it \rfckw{should} carry a timestamp and be
re-evaluated when the image, configuration, or control it relied on changes, so a
suppression cannot outlive the condition that justified it.

\section{Determinism, governance, and defensibility}
\begin{itemize}[leftmargin=1.4em]
\item \textbf{Determinism.} Given the same (CVE vector, archetype, VER axes), the
  method yields the same PAIN and deadline on every run and for every analyst.
\item \textbf{Two human inputs only.} An asset needs exactly two tags
  (\code{asset-archetype}, \code{multi-agency}); everything else is derived.
\item \textbf{Governed calibration.} The word thresholds
  (Eq.~\eqref{eq:word}) and the EPSS threshold are the only tunable knobs and
  \rfckw{must} live in governed configuration, not in ad-hoc cluster state.
\item \textbf{Auditability.} Each finding can carry the full derivation: the
  resolved archetype and its source, CR/IR/AR, $S$, the word, the scope, and the
  selected matrix cell.
\end{itemize}

\section{Security Considerations}
\label{sec:seccon}
The method trusts the type-tags on each system: if someone mislabels a critical
system as trivial, its flaws will look smaller than they really are. Its integrity
therefore depends on the trustworthiness of two inputs --- the asset archetype tags
and the VER axes (EPSS / KEV membership / reachability). Tag spoofing that
\emph{lowers} an archetype is a downgrade attack; the fail-safe default and governed
calibration mitigate accidental under-classification, but tag assignment
\rfckw{should} be controlled like any other security setting. Finally, the method
does not detect vulnerabilities; it prioritizes the output of a scanner and inherits
that scanner's coverage and accuracy.

\section{Conclusion}
FedRAMP Rev5 VDR/VER specifies the vocabulary and the axes of vulnerability
prioritization but leaves the classifier unspecified. That classifier need not be
invented: the CVSS Environmental metric group already encodes ``re-weight impact by
what is at stake,'' which is precisely the agency-impact question. By assigning each
asset its CR/IR/AR --- via archetypes or any other systematic scheme the CSP owns
--- and evaluating the two VER axes, a provider obtains a deterministic, auditable,
and standards-grounded PAIN and remediation deadline for every finding. We offer the
example archetype catalog not as a standard but as an existence proof; we encourage
each CSP to publish its own.

A companion proposal, \emph{A Finer Exploitability Gradation (VLEV) for FedRAMP
VDR/VER}, separately explores tightening the exploitability axis. It is kept out of
this memo deliberately: this memo's scope is meeting FedRAMP's requirements as they
stand today, not changing them. A second companion, \emph{A CycloneDX VEX Profile for
FedRAMP Rev5 VDR/VER Disposition and Response}, specifies the machine-readable format
for the disposition step of Section~\ref{sec:vex}.

\appendix
\section{Example archetype catalog}
\label{app:catalog}
Illustrative only (see Section~\ref{sec:archetypes}).

\begin{table}[H]
\centering
\renewcommand{\arraystretch}{1.2}
\footnotesize
\setlength{\tabcolsep}{5pt}
\begin{tabular}{l l c c c >{\raggedright\arraybackslash}p{0.34\textwidth}}
\toprule
\textbf{Archetype} & \textbf{Lens} & \textbf{CR} & \textbf{IR} & \textbf{AR} & \textbf{Typical members} \\
\midrule
\code{cicd-pipeline}      & control & H & H & H & build/deploy runners, artifact signing, registries \\
\code{orchestrator}       & control & H & H & H & control plane, etcd, scheduler, coordination \\
\code{config-actuation}   & control & H & H & H & IaC/GitOps, schema registry, admin/migration \\
\code{identity-secrets}   & control & H & H & H & IdP/SSO, KMS, secrets managers, session stores \\
\code{security-tooling}   & control & H & H & M & scanners, SIEM, EDR, runtime security \\
\code{change-record}      & control & M & M & M & ITSM/ticketing (record only) \\
\code{platform-foundation}& control & L & H & H & DNS, NTP, service discovery, L4 internal LBs (metadata only) \\
\code{data-sensitive}     & data    & H & H & H & PII/CUI datastores \\
\code{data-backbone}      & data    & H & H & H & queues, brokers, the system-of-record DB \\
\code{app-tier}           & data    & M & M & H & stateless services, APIs, UIs, caches \\
\code{batch-analytics}    & data    & M & M & L & ETL, reporting, analytics jobs \\
\code{public-edge}        & data    & L & L & H & load balancers, public web, ingress \\
\code{internal-tooling}   & data    & L & L & L & dashboards, metrics/log agents \\
\code{dev-test}           & data    & L & L & L & non-production \\
\midrule
\code{unclassified}       & ---     & H & H & H & fail-safe default for untagged assets \\
\bottomrule
\end{tabular}
\end{table}

\section{Remediation deadline matrices}
\label{app:matrix}
Days; $0.5 = 12$ hours. Columns in order LEV+IRV / LEV+NIRV / NLEV. PAIN-1 has no
FedRAMP deadline (any class). The day counts below are reproduced \emph{verbatim}
from FedRAMP's published \textbf{VDR-TFR-PVR} table;\footnote{FedRAMP,
\emph{Vulnerability Detection and Response (VDR)}, Rev5 provider rule, requirement
\texttt{VDR-TFR-PVR}. Generated markdown pinned at \vdrcite{} (commit of 2026-06-24);
accessed 2026-06-27. Mandated by CISA BOD~26-04; FedRAMP may revise the values.}
Classes~A and~B share one schedule. The only provider-side choice here is which
Certification Class the offering commits to --- not the numbers.

\begin{table}[H]
\centering
\footnotesize
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{l ccc @{\hspace{2em}} ccc @{\hspace{2em}} ccc}
\toprule
& \multicolumn{3}{c}{\textbf{Class A / B}} & \multicolumn{3}{c}{\textbf{Class C}} & \multicolumn{3}{c}{\textbf{Class D}} \\
\cmidrule(lr){2-4}\cmidrule(lr){5-7}\cmidrule(lr){8-10}
\textbf{PAIN} & L+I & L+N & NLEV & L+I & L+N & NLEV & L+I & L+N & NLEV \\
\midrule
N5 & 4  & 8   & 32  & 2  & 4   & 16  & 0.5 & 1  & 8  \\
N4 & 8  & 32  & 64  & 4  & 8   & 64  & 2   & 8  & 32 \\
N3 & 32 & 64  & 192 & 16 & 32  & 128 & 8   & 16 & 64 \\
N2 & 96 & 160 & 192 & 48 & 128 & 192 & 24  & 96 & 192 \\
\bottomrule
\end{tabular}

\vspace{0.5em}
\begin{minipage}{0.95\textwidth}
\footnotesize
\emph{Note:} Values are days measured from \emph{completed evaluation}
(\texttt{VDR-TFR-PVR} runs its timeframes from evaluation, not detection) within
which the finding \rfckw{should} be remediated, fully mitigated, or at least
partially mitigated to a lower PAIN rating; $0.5$ represents 12 hours.
\end{minipage}
\end{table}

\section{Explicit design opinions (calibratable choices)}
\label{app:opinions}
Everything in this section is a \emph{judgment call} we made --- not something
FedRAMP or CVSS dictates. We collect them in one place so reviewers can challenge
each one, and so any adopter knows exactly which dials are theirs to turn. None of
these is load-bearing for the \emph{method}; they are the parameters the method runs
on. Each item below names the value we use, why, and the fact that an implementation
\rfckw{may} change it.

\begin{enumerate}[leftmargin=1.7em]
\item \textbf{PAIN word thresholds $(\theta_1,\theta_2,\theta_3) = (0.25,\,0.55,\,0.80)$.}
  The cut points that turn the impact scalar $S$ into
  Minimal/Narrow/Disruptive/Debilitating (Eq.~\ref{eq:word}). This is the single most
  consequential opinion --- it sets where a finding crosses an N-level boundary. The
  values should reproduce senior-analyst triage on a back-tested sample; an adopter
  \rfckw{should} re-calibrate against its own rated findings and \rfckw{must} treat
  the result as governed configuration.
\item \textbf{EPSS Likely-Exploitable threshold $\theta_{\text{epss}} = 0.70$.}
  The EPSS probability at or above which a finding is treated as LEV
  (Section~\ref{sec:remediation}). Higher means fewer findings on the fast track;
  FedRAMP leaves the framework to the provider.
\item \textbf{LEV is a union, not a blend.} A finding is Likely Exploitable if
  EPSS $\ge \theta_{\text{epss}}$, \emph{or} it is listed in the CISA KEV catalog,
  \emph{or} it is internet-reachable via direct exposure and its vector permits
  unauthenticated automation (\code{AV:N/PR:N/UI:N}) --- whichever fires, with no
  weighting among them. The third disjunct is not provider discretion: it
  implements the \code{FRD-LEV} floor note that ``any vulnerability that an
  automated unauthenticated system can exploit over the internet is a likely
  exploitable vulnerability.''
\item \textbf{Impact-only severity, normalized by $0.915$.} We take CVSS's
  \emph{Modified Impact Sub-Score} and rescale it to $[0,1]$ (Eq.~\ref{eq:s}) rather
  than computing the full 0--10 environmental score. PAIN is about \emph{impact on the
  asset}, so we intentionally drop the exploitability half of the base score (and the
  $6.42$ coefficient) from the severity term; exploitability re-enters separately as
  the LEV/IRV remediation columns.
\item \textbf{CISA Vulnrichment is deliberately excluded.} Two CISA Vulnrichment
  SSVC values are natural candidate inputs: the exploitation state (as a
  LEV input) and the technical impact (as an impact floor). We adopt neither. The
  exploitation role is served by KEV membership --- the instrument BOD~26-04
  itself names --- and a technical-impact floor is rejected because an opaque,
  externally-fixed impact floor cannot coexist with the evidence-based
  Modified~C/I/A mitigation lever of Section~\ref{sec:ladder}: when a fixed floor
  and cited local evidence disagree, there is no principled precedence between
  them, and the Environmental layer already re-weights impact per deployment. EPSS
  and the KEV catalog remain the two governed threat inputs. An adopter \rfckw{may}
  reintroduce enrichment sources, at its own risk of exactly that conflict.
\item \textbf{Modifications lower High to Low, never to None; evidence scales with
  the PAIN reduced.} The mitigation lever of Section~\ref{sec:ladder} lets cited,
  signed evidence lower a Modified impact dimension, but never to None: a claim of
  \emph{zero} impact is an applicability claim and must travel as a VEX disposition
  (Section~\ref{sec:vex}), where it is separately governed, auditable, and
  reversible. And the evidence bar scales with the stakes --- lowering an N5 to an
  N4 demands stronger corroboration than lowering an N2 to an N1, mirroring the
  disposition guardrails. Both rules are ours, not FedRAMP's; an adopter tightening
  or relaxing them \rfckw{should} do so in governed configuration.
\item \textbf{Scope is purely hierarchical.} An asset is multi-agency only if the
  asset itself, a containing boundary, or the provider default is tagged so
  (most-specific wins); there is no automatic per-archetype escalation. (The
  tempting shortcut --- auto-escalating ``shared-infrastructure'' archetypes to
  multi-agency --- is rejected in favor of explicit, operator-controlled tagging.)
\item \textbf{``Archetype'', not ``type''.} We classify assets by their
  \emph{role/usage pattern}, not their intrinsic kind, and the label reflects that:
  the same software lands in different archetypes by how it is used (an in-memory
  store is \code{app-tier} as a cache, \code{identity-secrets} as a session store,
  \code{data-backbone} as a broker). ``Type'' would wrongly suggest an intrinsic
  property of the software and collides with overloaded terms (e.g.\ Kubernetes
  Service \code{type}); ``archetype'' names the pattern an asset matches in the
  estate. An adopter \rfckw{may} rename the label, but the role-not-kind semantics
  should be preserved.
\item \textbf{Fail-safe defaults to conservative High.} An unclassified asset defaults
  to CR/IR/AR all High so it scores loudly, rather than to a low default. ``Unknown''
  is treated as serious until classified.
\item \textbf{CVSS v3.1, not v4.0.} We build on v3.1's closed-form environmental
  formula because it provides a Modified Impact Sub-Score (Eq.~\ref{eq:isc}) with
  explicit CR/IR/AR multipliers. CVSS v4.0 retains CR/IR/AR but feeds them into a
  lookup-based MacroVector rather than the v3.1 formula, and it splits impact into
  Vulnerable-System (\code{VC/VI/VA}) and Subsequent-System (\code{SC/SI/SA})
  metrics. A v4.0-only finding can be approximated by mapping \code{VC/VI/VA} onto
  $C/I/A$, at the cost of ignoring Subsequent-System impact; while NVD remains
  v3.1-dominant, an implementation \rfckw{should} prefer the v3.1 vector when both
  are available.
\item \textbf{The archetype catalog and its CR/IR/AR values
  (Appendix~\ref{app:catalog}).} Illustrative only --- each CSP \rfckw{should} derive
  its own from its system categorization. A 3PAO \rfckw{may} take on the role of 
  validating these mappings during an assessment; however, the CSP remains 
  ultimately responsible if an asset is undertagged, a vulnerability is not 
  remediated in time, and it is subsequently exploited (just as in the 
  commercial world).
\item \textbf{The remediation deadline values (Appendix~\ref{app:matrix}) are
  \emph{not} ours to calibrate.} They are reproduced verbatim from FedRAMP's published
  \texttt{VDR-TFR-PVR} table (mandated by CISA BOD~26-04; see Section~\ref{sec:bod}).
  The only provider-side choice is which Certification Class (A--D) the offering
  commits to, selecting the column block; the day counts themselves are fixed by
  FedRAMP. Listed here only to mark the boundary between what we chose and what the
  rule dictates.
\end{enumerate}

\section{Document version history}
\label{app:history}
This memo is a living document. Table~\ref{tab:history} tracks its version history.

\begin{table}[H]
\centering
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lll>{\raggedright\arraybackslash}p{0.55\textwidth}}
\toprule
\textbf{Version} & \textbf{Date} & \textbf{Author} & \textbf{Summary of Changes} \\
\midrule
1.0 & June 28, 2026 & M. Venne & Initial publication. \\
1.2 & June 29, 2026 & M. Venne & Various readability updates. \\
\bottomrule
\end{tabular}
\caption{Document version history.}
\label{tab:history}
\end{table}

\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}
While this paper has a single official author, the underlying work, refinement,
and practical validation were a collaborative effort. It would not have been
possible without the generous help, critical feedback, and dedicated support of
various other team members at stackArmor who helped make it possible.

\section*{Normative sources}
\addcontentsline{toc}{section}{Normative sources}
\begin{itemize}[leftmargin=3.4em]
\item[{[VDR]}] FedRAMP, \emph{Vulnerability Detection and Response (VDR)}, Rev5.
  Requirements cited: \texttt{VDR-TFR-PVR}, \texttt{VDR-TFR-KEV},
  \texttt{VDR-TFR-PDD/PCD/PSD}. Pinned permalink: \vdrcite{} (commit of 2026-06-24;
  generated from upstream \texttt{fedramp/rules}). Accessed 2026-06-27.
\item[{[VER]}] FedRAMP, \emph{Vulnerability Evaluation and Reporting (VER)}, Rev5.
  Requirements cited: \texttt{VER-EVA-EIR}, \texttt{VER-EVA-EFP},
  \texttt{VER-EVA-AIA}, \texttt{VER-RPT-VDT}, \texttt{VER-TFR-MAV},
  \texttt{VER-RPT-AVI}. Pinned permalink: \vercite{} (commit of 2026-06-24).
  Accessed 2026-07-01.
\item[{[DEF]}] FedRAMP, \emph{FedRAMP Definitions}. Definitions cited:
  \texttt{FRD-IRV}, \texttt{FRD-LEV}. Pinned permalink: \defcite{} (commit of
  2026-06-24). Accessed 2026-07-01.
\item[{[EGR]}] M.~Venne, \emph{An Exclusion-Gated Method for Determining Internet
  Reachability of Vulnerabilities under the FedRAMP VDR/VER Rules}, stackArmor,
  July 2026. Companion paper specifying the reachability determination used here.
\item[{[BOD]}] CISA, \emph{Binding Operational Directive 26-04}, 2026-06-10.
\item[{[KEV]}] CISA, \emph{Known Exploited Vulnerabilities (KEV) Catalog}. LEV
  input; \texttt{VDR-TFR-KEV} schedule.
\item[{[CVSS]}] FIRST, \emph{CVSS v3.1 Specification} --- Environmental metric group
  and Modified Impact Sub-Score.
\item[{[EPSS]}] FIRST, \emph{Exploit Prediction Scoring System}.
\end{itemize}

\end{document}
