\documentclass[11pt]{article}

\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}
\usepackage{float}
\usepackage{xcolor}
\usepackage{listings}
\usepackage[hidelinks]{hyperref}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\rfckw}[1]{\textsc{#1}}

\definecolor{lstbg}{HTML}{F6F8FA}
\definecolor{lstrule}{HTML}{D0D7DE}
\lstdefinestyle{json}{
  basicstyle=\ttfamily\scriptsize,
  backgroundcolor=\color{lstbg},
  frame=single, rulecolor=\color{lstrule},
  framesep=6pt, xleftmargin=6pt, xrightmargin=6pt,
  breaklines=true, breakatwhitespace=true,
  columns=fullflexible, keepspaces=true,
  showstringspaces=false, aboveskip=10pt, belowskip=6pt,
}

\title{\bfseries A CycloneDX VEX Profile for\\ FedRAMP Rev5 VDR/VER Disposition and Response}
\author{Matthew Venne \\ \small Chief Technology Officer, stackArmor}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
This is a companion to \emph{A Deterministic, CVSS--Environmental Method for FedRAMP
Rev5 VDR/VER Vulnerability Prioritization}, which assigns every detected finding a
Potential Agency Impact rating (PAIN, N1--N5) and a remediation deadline by formula.
That scoring is deterministic and computable \emph{a priori}. Deciding whether a
finding is \emph{truly} exploitable in the running environment --- and, when a fix will
miss the deadline, attesting the compensating response --- is a separate,
\emph{a posteriori} step that can only follow detection. FedRAMP's Vulnerability
Evaluation Requirements (VER) make that step normative but name no format. This
proposal recommends \textbf{CycloneDX VEX} as the structured, machine-readable, signable
carrier for it, chosen for the fit of its disposition and response \emph{vocabulary} ---
not because it carries scores; it does not. We give the field mapping to VER, a thin
worked example, an OCI-native distribution pattern, and an optional crosswalk to CISA's
status-justification labels for downstream interoperability. This document is an
\emph{implementation profile}: it shows one defensible way to satisfy existing VER
requirements; it does not propose changing them. The key words \rfckw{must},
\rfckw{should}, and \rfckw{may} are to be read as in RFC~2119.
\end{abstract}

\section{Where disposition sits in the lifecycle}
The method memo computes PAIN and a deadline for \emph{every} finding a scanner emits.
But a scan hit is a candidate, not a confirmed exposure: the flagged code may be
absent, may never execute, may require a configuration nobody enabled, or may sit
behind a control that already neutralizes it. Resolving which is the \emph{disposition}
step.

Disposition is an \emph{a posteriori} determination. The factors that decide it ---
reachability, configuration, surrounding mitigations --- are specific to a given
vulnerability on a given deployment and cannot be enumerated before that vulnerability
is known. The analysis therefore necessarily follows detection, acutely so for novel
vulnerabilities, where no prior due-diligence could exist. This is a statement about
\emph{timing}, not about who or what performs it: the determination \rfckw{may} be fully
automated or AI/agentic, but it cannot be precomputed.

Because VER's clocks are short, the determination must be captured in a form that
machines can produce, exchange, sign, and act on. A disposition resolves a finding onto
one of two tracks:
\begin{itemize}[leftmargin=1.5em,nosep,topsep=3pt]
\item \textbf{Not truly affected} --- the finding is suppressed, with a machine-readable
  justification recording \emph{why} (code absent, unreachable, configuration-gated, or
  already mitigated).
\item \textbf{Affected but remediation will be late} --- the finding is accepted within
  process, with an attested mitigation and a planned response, rather than silently
  breaching the deadline.
\end{itemize}

\section{Why CycloneDX: the vocabulary fits the lifecycle}
FedRAMP is not bound to any single VEX format, and VER specifies \emph{fields and
content}, not a schema (Section~\ref{sec:vermap}). The provider therefore chooses the
carrier --- the same latitude FedRAMP itself exercised in substituting PAIN for SSVC.
On that basis the deciding factor is \emph{expressiveness}: which format's disposition
vocabulary most precisely captures the real outcomes above. CycloneDX's \code{analysis}
object is the strongest fit. The same granularity pays off when disposition is
automated: because each justification names a distinct precondition, an automated or
AI/agentic analyzer can map each evidence source one-to-one to a justification, where a
coarser vocabulary would collapse several distinct findings onto a single label.

\begin{table}[H]
\centering
\footnotesize
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.30\textwidth} >{\raggedright\arraybackslash}p{0.62\textwidth}}
\toprule
\textbf{CycloneDX field} & \textbf{Values (and what they capture)} \\
\midrule
\code{analysis.state} & \code{resolved}, \code{resolved\_with\_pedigree},
  \code{exploitable}, \code{in\_triage}, \code{false\_positive},
  \code{not\_affected} --- note a first-class \code{false\_positive} and an
  \code{in\_triage} (work-in-progress) state. \\
\code{analysis.justification} & \code{code\_not\_present}, \code{code\_not\_reachable},
  \code{requires\_configuration}, \code{requires\_dependency},
  \code{requires\_environment}, \code{protected\_by\_compiler},
  \code{protected\_at\_runtime}, \code{protected\_at\_perimeter},
  \code{protected\_by\_mitigating\_control} --- nine reasons, naming configuration,
  reachability, and perimeter/runtime controls explicitly. \\
\code{analysis.response} & \code{can\_not\_fix}, \code{will\_not\_fix}, \code{update},
  \code{rollback}, \code{workaround\_available} --- the planned action, which the
  late-remediation track needs and lighter formats lack. \\
\code{analysis.detail} & free text for human-readable context and evidence. \\
\bottomrule
\end{tabular}
\caption{CycloneDX's \code{analysis} block: states, justifications, and responses.}
\end{table}

The operational cases a CSP actually encounters map onto native values without
contortion:
\begin{table}[H]
\centering
\footnotesize
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.42\textwidth} >{\raggedright\arraybackslash}p{0.50\textwidth}}
\toprule
\textbf{Real-world disposition} & \textbf{CycloneDX expression} \\
\midrule
Scanner false positive & \code{state: false\_positive} \\
Vulnerable code absent / removed & \code{not\_affected} + \code{code\_not\_present} \\
Vulnerable path never executes & \code{not\_affected} + \code{code\_not\_reachable} \\
Needs an unset configuration & \code{not\_affected} + \code{requires\_configuration} \\
Neutralized at the edge (WAF/IAP) & \code{not\_affected} + \code{protected\_at\_perimeter} \\
Neutralized at runtime / by a control & \code{not\_affected} + \code{protected\_at\_runtime} / \code{protected\_by\_mitigating\_control} \\
Affected; fix will miss deadline & \code{exploitable} + \code{response:[workaround\_available, update]} + \code{detail} \\
Under evaluation & \code{in\_triage} \\
\bottomrule
\end{tabular}
\caption{Operational dispositions and their native CycloneDX encodings.}
\label{tab:cases}
\end{table}

\section{Mapping to FedRAMP VER}
\label{sec:vermap}
VER requires the disposition step and a machine-readable report, in each case by
function rather than by named format. CycloneDX satisfies each clause:

\begin{table}[H]
\centering
\footnotesize
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.30\textwidth} >{\raggedright\arraybackslash}p{0.62\textwidth}}
\toprule
\textbf{VER requirement} & \textbf{CycloneDX mechanism} \\
\midrule
\code{VER-EVA-EFP} (evaluate false positives) & \code{analysis.state = false\_positive} \\
\code{VER-EVA-AIA} (assume exploitable absent evidence) & a VEX statement \emph{is} the
  ``evidence otherwise''; its absence leaves the assume-exploitable default in force \\
\code{VER-RPT-VDT} (per-finding disposition) & \code{analysis.state} +
  \code{justification} + \code{response} keyed to the CVE and component \\
\code{VER-TFR-MAV} / \code{VER-RPT-AVI} (accepted vulns) & \code{state = exploitable}
  with \code{response} and \code{detail} recording the acceptance and mitigation \\
\code{VER-TFR-MRH} (machine-readable JSON history) & CycloneDX is JSON; statements
  accrete over time with timestamps \\
\bottomrule
\end{tabular}
\caption{FedRAMP VER clauses and the CycloneDX fields that satisfy them.}
\end{table}

\section{A thin attestation: what it carries, and what it does not}
The disposition artifact stays deliberately small. It carries only what identifies the
finding and records the judgment:
\begin{itemize}[leftmargin=1.5em,nosep,topsep=3pt]
\item the vulnerability \code{id} and its \code{source};
\item the affected component, by \code{affects[].ref} (a package URL / \code{bom-ref});
\item the \code{analysis} block --- \code{state}, \code{justification},
  \code{response}, \code{detail};
\item provenance --- author, timestamp, and a signature over the artifact.
\end{itemize}

It deliberately \emph{omits} PAIN, the LEV/IRV axes, the Certification Class, and the
CVSS vector. Those are computed deterministically by the method and live in the
provider's scoring report; duplicating them into the disposition artifact would invite
drift and tempt a reader to treat an attestation as a place to re-score. The two
records are joined on a stable key --- the \textbf{CVE identifier plus the component
reference} --- so a reporting layer can attach each disposition to the finding it
governs and recover the retained PAIN for audit. The disposition changes a finding's
\emph{handling}; it never changes its score.

\section{Worked example}
A single signed CycloneDX document can carry several statements. Below, one finding is
dispositioned \code{not\_affected} (unreachable code) and one is \code{exploitable} with
an attested perimeter mitigation because the vendor fix will arrive after the deadline.
Note the absence of any score, rating, or PAIN field.

\begin{lstlisting}[style=json]
{
  "bomFormat": "CycloneDX",
  "specVersion": "1.6",
  "version": 1,
  "metadata": {
    "timestamp": "2026-06-28T14:00:00Z",
    "authors": [{ "name": "CSP Security Team" }],
    "component": {
      "type": "application",
      "name": "payments-api",
      "bom-ref": "payments-api@2.4.1"
    }
  },
  "vulnerabilities": [
    {
      "id": "CVE-2026-10101",
      "source": { "name": "NVD",
        "url": "https://nvd.nist.gov/vuln/detail/CVE-2026-10101" },
      "analysis": {
        "state": "not_affected",
        "justification": "code_not_reachable",
        "detail": "Vulnerable parser entrypoint is never invoked; only the
                   signing helper from this library is linked."
      },
      "affects": [{ "ref": "pkg:golang/example.com/parser@1.2.3" }]
    },
    {
      "id": "CVE-2026-20202",
      "source": { "name": "NVD",
        "url": "https://nvd.nist.gov/vuln/detail/CVE-2026-20202" },
      "analysis": {
        "state": "exploitable",
        "response": ["workaround_available", "update"],
        "detail": "Vendor fix ETA exceeds the VER deadline. Perimeter WAF
                   virtual patch deployed and egress restricted; tracked for
                   update on vendor release."
      },
      "affects": [{ "ref": "pkg:deb/debian/libxyz@2.1.0-1" }]
    }
  ]
}
\end{lstlisting}

\section{Distribution and retrieval over OCI}
For containerized workloads the disposition artifact \rfckw{should} travel with the
image it describes. CycloneDX documents have a defined media type and can be stored as
OCI artifacts in the same registry as the image and associated with it through the
registry's referrers/attestation mechanism. A scanner that already pulls an image to
detect vulnerabilities can, in the same pass, resolve the image's attached VEX and
apply the dispositions automatically --- so the \emph{a posteriori} judgment, once made
and signed, is rediscovered and reused on every subsequent scan without re-litigation.
This keeps disposition co-located with the artifact under evaluation, versioned with it,
and verifiable by signature, and folds it natively into the existing detection workflow
rather than bolting on a side channel.

\section{Optional interoperability: CycloneDX to CISA labels}
A FedRAMP-facing report needs nothing beyond the above. Some downstream consumers ---
agency tooling, or a CSAF advisory pipeline --- speak CISA's narrower
status-justification vocabulary instead. CISA, like CycloneDX, separates the two axes:
each CycloneDX \code{state} coarsens onto a CISA \emph{status} and each
\code{justification} onto a CISA \emph{justification label}, in both cases without loss
of safety (the mapping only ever generalizes; it never invents a stronger claim). This
crosswalk is \rfckw{optional} and exists solely for interchange.

\begin{table}[H]
\centering
\footnotesize
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.40\textwidth} >{\raggedright\arraybackslash}p{0.52\textwidth}}
\toprule
\textbf{CycloneDX justification / state} & \textbf{CISA label (coarsened)} \\
\midrule
\code{code\_not\_present} & \code{vulnerable\_code\_not\_present} \\
\code{code\_not\_reachable} & \code{vulnerable\_code\_not\_in\_execute\_path} \\
\code{requires\_configuration} & \code{vulnerable\_code\_not\_in\_execute\_path} \\
\code{requires\_dependency} & \code{vulnerable\_code\_not\_in\_execute\_path} \\
\code{requires\_environment} & \code{vulnerable\_code\_not\_in\_execute\_path} \\
\code{protected\_by\_compiler} & \code{inline\_mitigations\_already\_exist} \\
\code{protected\_at\_runtime} & \code{inline\_mitigations\_already\_exist} \\
\code{protected\_at\_perimeter} & \code{inline\_mitigations\_already\_exist} \\
\code{protected\_by\_mitigating\_control} & \code{inline\_mitigations\_already\_exist} \\
\midrule
\code{state: not\_affected} & \code{not\_affected} (paired with the justification above as the CISA label) \\
\code{state: false\_positive} & \code{not\_affected} (\code{component\_not\_present} where the component is truly absent) \\
\code{state: in\_triage} & \code{under\_investigation} \\
\code{state: exploitable} & \code{affected} \\
\code{state: resolved} / \code{resolved\_with\_pedigree} & \code{fixed} \\
\bottomrule
\end{tabular}
\caption{Optional coarsening of CycloneDX dispositions onto CISA's labels/statuses.}
\end{table}

\section{Standing and adoption}
CycloneDX introduced its vulnerability and VEX capabilities in version~1.4
(January~2022), ahead of CISA's status-justification guidance (June~2022); its richer
vocabulary is thus an independent, earlier design rather than a divergence from a later
one. CycloneDX is now an Ecma International standard (ECMA-424). Adoption is additive:
the disposition layer sits on top of the unchanged scoring pipeline of the method memo,
joined by CVE and component, and a provider \rfckw{may} emit the optional CISA crosswalk
for consumers that require it. As with the method's archetype catalog, this profile is
offered as a workable existence proof, not as a new FedRAMP standard --- the choice of
carrier remains the provider's.

\end{document}
